Legacy Data Confounds Genomics Studies

### 2019-04-03 legacy.tex
update : greyed out text and minor text changes

To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use genotypes as the independent variable. 
Here we perform a ``reverse GWAS'', in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable.
We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}A).
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.

@@

\begin{table}[h]
\begin{tabular}{l l l r r r}
  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)}      \\ \hline
28654678 & PLoS One & *rs201761909 & 5.7 & 78.11\\
	& 	&  	*rs201130852 & 5.05 & 72.28\\
	& 	&  	*rs201255786 & 5.7 & 68.97\\
	& 	&  	*rs200655768 & 6.52 & 66.67\\
	& 	&  	*rs184202621 & 5.52 & 60.45\\
	& 	& 	*rs80274284 & 6 & 56.15\\
	& 	& 	*rs200699422 & 5.3 & 7.43\\
23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & *rs6057648 & 5.4 & 20.5\\
28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & *rs201471471 & 6.52 & 7.87\\
28270201 & \multicolumn{1}{p{3cm}}{\raggedright Genome Medicine} & rs453755 & 7.52 & 5.29\\
23023329 & \multicolumn{1}{p{3cm}}{\raggedright Nature Genetics} & rs103294 & *15.3 & 4.32\\
28334899 & \multicolumn{1}{p{3cm}}{\raggedright Human Molecular Genetics} & rs103294 & *29.3 & 4.32\\
28240269 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs103294 & *72.7 & 4.32\\
27863252 & \multicolumn{1}{p{3cm}}{\raggedright Cell\\} & rs3794738 & *13.15 & 3.73\\
29534301 & \multicolumn{1}{p{3cm}}{\raggedright Hepatology\\} & rs9273062 & *9.7 & 3.36\\
21386085 & \multicolumn{1}{p{3cm}}{\raggedright Diabetes\\}  & rs301 & *10.52 & 3.02\\
26830138 & \multicolumn{1}{p{3cm}}{\raggedright Molecular Psychiatry} & rs77894924 & 6.7 & 2.77\\
29617998 & \multicolumn{1}{p{3cm}}{\raggedright Human Molecular Genetics} & rs4963156 & *22.4 & 2.52\\
28698626 & \multicolumn{1}{p{3cm}}{\raggedright Scientific Reports} & rs11015915 & 5.05 & 2.45\\
26974007 & \multicolumn{1}{p{3cm}}{\raggedright Nature Genetics} & rs3124998 & *8.05 & 2.33\\
26634245 & \multicolumn{1}{p{3cm}}{\raggedright BMC Genetics} & rs451000 & 6 & 2.28\\
 	& 	& 	rs443874 & 5.3 & 2.26\\
	& 	&	rs400942 & 6 & 2.2\\
26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & *rs60136336 & 5.7 & 2.25\\

25918132 & \multicolumn{1}{p{3cm}}{\raggedright Toxicological Sciences} & rs76780579 & 6 & 2.09\\

 \hline
\end{tabular}
\caption{A list of recent publications that reported suspicious variants as close to or above the genome-wide significant threshold. The variants reaching genome wide significance have a star ( * ). Note : with a FDR ($\alpha = 0.01$) p-value adjustment, values of $ p > -\log_{10}(\alpha)$ are deemed statistically significant. }
\label{gwasTable}
\end{table}

@@

\subsection{Recommendations}
While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
In order to avoid spurious results, the most conservative approach would be to remove or resequence all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$. 
We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{}).
Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p < 10^{-2}$ in the reverse GWAS. (The supplementary material contains a list of variants statistically associated to $Q$).

@@

\section{Methods}
\subsection{Metadata}
The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
Average quality of mapped bases $Q$ per sample was obtained from the BAS files associated with each alignment file. 
Each BAS file has metadata regarding each sequencing event for each sample. 
If a sample was sequenced more than once, we took the average of the each $Q$ score from each sequencing instance. 
The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files.  
This file also has multiple entries per sample, however, we were unable to match the individual sequencing runs between the bas files and the index file, which lead us to take the average of the $Q$ scores and only kept the earliest sequencing date per sample. 
The dates of the sequencing are only used to plot Figure.

@@

\todo{link to my compiled metadata file here}

@@

%Examples of models that are widely used to account population structure include the Balding-Nichols model \citep{balding1995method}, and the Pritchard- Stephens-Donnelly model \citep{pritchard2000inference}. These and several other similar models used in GWAS studies can be understood in terms of the following matrix factorization. 
%\begin{align}
%\mathbf{L }= \mathbf{AH}
%\end{align} 
%where the $i^\text{th}$ column ($\boldsymbol{h}^{(i)}$) of the $K \times I$ matrix $\mathbf{H}$ encodes the population structure of the $i$'th individual and the $s$'th row of the $S \times K$ matrix $\mathbf{A}$ determines how that structure is manifested in SNP $s$. When Hardy-Weinberg equilibrium holds, observed genotype can be assumed to be generated by the following Binomial model.
%\begin{align} \label{lfa}
%y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big) 
%\end{align} 
%for $s=1\hdots S$ and $i=i,\cdots, I$, where $y_{si} \in \{0,1,2\}$ and $logit(\pi_{si})$ is the $(s,i)$ element of the matrix $\mathbf{L}$ such that  $\pi_{si}$ is the individual-specific allele frequency.
%
%To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Association Test  (GCAT) proposed by \citep{song2015testing} The GCAT is a regression approach that assumes the following model.
%\begin{align}\label{gcat}
% y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
%\end{align} 
%for $s=1\hdots S$ and $i=i,\cdots, I$ \luke{Plug in the values of S and I} and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$. The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and plugged in the model. We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs. The matrix \luke{Describe matrix}

@@

\subsection{Genome wide association study}
We ran a simple logistic regression independently for each genomic locus $s$ using the R glm2 package\citep{RDevelopmentCoreTeam2016}. The model we fitted to each subset of data can be expressed as such.
\begin{align*}
%{logit}{E(Y_{s'i})} &= \beta_{0}  + \beta_{1}{PC1}_{i} + \beta_{2}{PC2}_{i} + \beta_{3}{PC3}_{i} + \beta_{4}{PC4}_{i} + \beta_{5}{PC5}_{i} + \beta_{6}Q_{i}
{logit}{E(Y_{s'i})} &= \beta_{0}  + \beta_{1}{Pop}_{i}  + \beta_{2}Q_{i}
\\
\textit{for}\quad i &= 1,\hdots, n
\end{align*}
Here $Y_{si}$ is a Bernoulli random variable denoting the genotype of individual $i$ at genomic locus $s$. 
%In the model, the regression coefficient $\beta_{1}{PC1}$ represents the coefficient of the first principal component, the slope coefficient $\beta_{6}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$. 
In the model, the regression coefficient $\beta_{1}{Pop}_{i}$ represents the coefficient of the categorical variable of Population, the slope coefficient $\beta_{6}$ denotes the association to average quality of mapped bases $Q$ to genotype $Y_{s}$. 
We run a total of $S$ regressions, where $S$ is the total number of genomic loci.

For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{2}^{(s)}=0$. That is, the average quality of mapped bases $Q$ variable is not associated with genotype $Y_{s}$.
In other words, the slope coefficient of the simple logistic regression is zero.
To test the null hypothesis, we used the chi-square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of adding $Q$ in the model. 
The deviance test statistic is approximately chi-square distributed with one degrees of freedom.

Given the large number of tests (of which there are $S$), the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values \citep{Benjamini2006}.
By using a nominal Type-I error rate $\alpha = 0.01$, a total of 24,390 variants were found to be statistically significance. See supplementary file {here}. \luke{make this file}

>>>>

To identify SNPs that are likely to reproduce poorly across cohorts without having access to a second cohort, we performed an association study in the JPT for SNPs that associate strongly with low $Q$ (Figure \ref{SFS}).
Traditionally, genome wide association studies use genotypes as the independent variable. 
Here we perform a ``reverse GWAS'' or Genotype-Conditional Association Test, in the sense that genotypes are now the dependent variable that we attempt to predict using the continuous variable $Q$ as the independent variable \citep{song2015testing}.
We use logistic regression of the genotypes on $Q$ and identify 587 SNPs with $p < 10^{-8}$ and 1034 SNPs with $ p < 10^{-6}$. 
While identifying putative low-quality SNPs to exclude, using a higher $p$-value threshold increases the stringency of the filtering (i.e., excluding SNPs with $ p < 10^{-6}$ is more stringent than excluding SNPS with $p < 10^{-8}$). 
The variants that are associated to $Q$ have an enrichment in *AC${\rightarrow}$*CC mutations, GA*${\rightarrow}$GG*, and GC*${\rightarrow}$GG* mutations (Figure \ref{SFS}A).
These three enrichments can be summarized as an excess of G**${\rightarrow}$GG* in individuals with low $Q$.

@@

\begin{table}[h]
\begin{tabular}{l l l r r r}
  {Pubmed ID}  & {Journal} & {rsID} & \multicolumn{1}{p{1cm}}{\centering GWAS \\ -$\log_{10} p$} & \multicolumn{1}{p{1.5cm}}{\centering $Q$ -$\log_{10} p$\\(adjusted)}      \\ \hline
28654678 & PLoS One & rs201761909 & 5.7 & 78.11\\
	& 	&  	rs201130852 & 5.05 & 72.28\\
	& 	&  	rs201255786 & 5.7 & 68.97\\
	& 	&  	rs200655768 & 6.52 & 66.67\\
	& 	&  	rs184202621 & 5.52 & 60.45\\
	& 	& 	rs80274284 & 6 & 56.15\\
	& 	& 	rs200699422 & 5.3 & 7.43\\
23527680 & \multicolumn{1}{p{3cm}}{\raggedright Twin Research and\\Human Genetics} & rs6057648 & 5.4 & 20.5\\
28928442 & \multicolumn{1}{p{3cm}}{\raggedright Nature Communications} & rs201471471 & 6.52 & 7.87\\
26053186 & \multicolumn{1}{p{3cm}}{\raggedright PLoS One} & rs60136336 & 5.7 & 2.25\\ \hline
\textcolor{gray}{28270201} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Genome Medicine}} & \textcolor{gray}{rs453755} & \textcolor{gray}{7.52} & \textcolor{gray}{5.29}\\
\textcolor{gray}{23023329} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*15.3} & \textcolor{gray}{4.32}}\\
\textcolor{gray}{28334899} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*29.3} & \textcolor{gray}{4.32}\\
\textcolor{gray}{28240269} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Communications}} & \textcolor{gray}{rs103294} & \textcolor{gray}{*72.7} & \textcolor{gray}{4.32}\\
\textcolor{gray}{27863252} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Cell}\\} & \textcolor{gray}{rs3794738} & \textcolor{gray}{*13.15} & \textcolor{gray}{3.73}\\
\textcolor{gray}{29534301} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Hepatology}\\} & \textcolor{gray}{rs9273062} & \textcolor{gray}{*9.7} & \textcolor{gray}{3.36}\\
\textcolor{gray}{21386085} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Diabetes}\\}  & \textcolor{gray}{rs301} & \textcolor{gray}{*10.52} & \textcolor{gray}{3.02}\\
\textcolor{gray}{26830138} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Molecular Psychiatry}} & \textcolor{gray}{rs77894924} & \textcolor{gray}{6.7} & \textcolor{gray}{2.77}\\
\textcolor{gray}{29617998} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Human Molecular Genetics}} & \textcolor{gray}{rs4963156} & \textcolor{gray}{*22.4} & \textcolor{gray}{2.52}\\
\textcolor{gray}{28698626} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Scientific Reports}} & \textcolor{gray}{rs11015915} & \textcolor{gray}{5.05} & \textcolor{gray}{2.45}\\
\textcolor{gray}{26974007} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Nature Genetics}} & \textcolor{gray}{rs3124998} & \textcolor{gray}{*8.05} & \textcolor{gray}{2.33}\\
\textcolor{gray}{26634245} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{BMC Genetics}} & \textcolor{gray}{rs451000} & \textcolor{gray}{6} & \textcolor{gray}{2.28}\\
 	& 	& 	\textcolor{gray}{rs443874} & \textcolor{gray}{5.3} & \textcolor{gray}{2.26}\\
	& 	&	\textcolor{gray}{rs400942} & \textcolor{gray}{6} & \textcolor{gray}{2.2}\\
\textcolor{gray}{25918132} & \multicolumn{1}{p{3cm}}{\raggedright \textcolor{gray}{Toxicological Sciences}} & \textcolor{gray}{rs76780579} & \textcolor{gray}{6} & \textcolor{gray}{2.09}\\

 \hline
\end{tabular}
\caption{A list of recent publications that reported suspicious variants as close to or above the genome-wide significant threshold. The variants reaching genome wide significance have a star ( * ). Note : with a FDR ($\alpha = 0.01$) p-value adjustment, values of $ p > -\log_{10}(\alpha)$ are deemed statistically significant. The black text colour indicates that this variant may be spurious, grey text colour indicates that these variants are likely correct but may be impacted by a calling bias.}
\label{gwasTable}
\end{table}

@@

\subsection{Recommendations}
While it may be difficult to determine the exact source of the bias in the data, we have been able to identify individuals and loci that do not meet quality standards expected from high-quality sequencing data.
In order to avoid spurious results, the most conservative approach would be to remove or resequence all individuals that do not meet the quality threshold as well as all the variants associated to low $Q$. 
We found that a cut-off of $Q = 30$ was sufficient to remove the most egregious spurious signals (Supplementary Figure \ref{MutSpect}).
Another approach, for researchers relying on imputation servers where reference panel individuals cannot be filtered out, is to remove sites that reach a significance below $ p > -\log_{10}(\alpha)$ in the reverse GWAS. (The supplementary material contains a list of variants statistically associated to $Q$).

@@

\section{Methods}
\subsection{Metadata}
The metadata used in this analysis was compiled from each of the index files from the 1000 Genomes file system. 
Average quality of mapped bases $Q$ per sample was obtained from the BAS files associated with each alignment file. 
Each BAS file has metadata regarding each sequencing event for each sample. 
If a sample was sequenced more than once, we took the average of the each $Q$ score from each sequencing instance. 
The submission dates and sequencing centres for each sample in the analysis was available in the sequence index files \ref{ProcessPlotData.Rmd}.

@@

@@

@@

\subsection{Individual-specific allele frequency}
Examples of models that are widely used to account population structure include the Balding-Nichols model \citep{balding1995method}, and the Pritchard- Stephens-Donnelly model \citep{pritchard2000inference}. 
These and several other similar models used in GWAS studies can be understood in terms of the following matrix factorization. 
\begin{align}
\mathbf{L }= \mathbf{AH}
\end{align} 
where the $i^\text{th}$ column ($\boldsymbol{h}^{(i)}$) of the $K \times I$ matrix $\mathbf{H}$ encodes the population structure of the $i^\text{th}$ individual and the $s^\text{th}$ row of the $S \times K$ matrix $\mathbf{A}$ determines how that structure is manifested in SNP $s$. 
When Hardy-Weinberg equilibrium holds, observed genotype can be assumed to be generated by the following Binomial model.
\begin{align} \label{lfa}
y_{si} \mid \pi_{si}  &\sim Binomial\big( 2, \pi_{si} \big) 
\end{align} 
for $s=1\hdots S$ and $i=i,\cdots, I$, where $y_{si} \in \{0,1,2\}$ and $logit(\pi_{si})$ is the $(s,i)$ element of the matrix $\mathbf{L}$ such that  $\pi_{si}$ is the individual-specific allele frequency.

To test whether quality is associated to genotype while adjusting for population structure, we performed the Genotype-Conditional Association Test  (GCAT) proposed by \citep{song2015testing}.
The GCAT is a regression approach that assumes the following model.
\begin{align}\label{gcat}
 y_{si} \mid q_i, \boldsymbol{h}^{(i)} &\sim Binomial\bigg( 2, logit^{-1}\Big( \sum_{k=0}^{K} a_{sk} h_{ki} + \beta_s q_i\Big) \bigg)
\end{align} 
for $s=1\hdots S$ and $i=i,\cdots, I$  ($S = 28,516,063$ and $I = 2,504$) and where $\hat{h}_{0i}=1$ so that $a_{s0}$ is the intercept term and $logit(\pi_{si})=\sum_{k=0}^{K} a_{sk} h_{ki}$. 
The vectors $\boldsymbol{h}^{i}$ of the matrix $\mathbf{H}$ are unobserved but can be estimated using Logistic Factor Analysis (LFA) \citep{song2015testing} and plugged in the model. 
We approximated the population structure using $K=4$ latent components from a subsampled genotype matrix consisting of $M$ SNPs. 
The matrix includes the sequence data from the 2,504 individuals from the 1kGP down-sampled to the positions from the \href{ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/supporting/hd_genotype_chip/ALL.chip.omni_broad_sanger_combined.20140818.snps.genotypes.vcf.gz}{Omni chip}
